A Fast kNN Algorithm Using Multiple Space-Filling Curves

### 2022-05-14 2022/LION/main.tex
change the size of the figure
delete one word

\keywords{Multicriterial Problems \and Global Optimization \and Machine Learning \and Logistic Regression.}
\end{abstract}
%
%
%
\section{Introduction}
Machine learning (ML) methods are powerful tools that are applied in many areas of research.
In particular, the ML methods are successfully used for solving complicated problems of computational mathematics. One of such problem classes to which ML can be applied is the class of multicriterial optimization (MCO) models. 
In this case, ML is used, as a rule, in combination with metaheuristic algorithms \cite{Ma2014,Subraveti2019,Talbi2015} that in the case of multiextremal criteria lose in efficiency to deterministic methods \cite{Kvasov2018,Sergeyev2018}.

@@

Any effective (Pareto-optimal) variant  in which it is impossible to reduce the values of all criteria $f_i (y)$, $1 \leq i \leq s$, at once by choosing another option   can be considered as a partial solution to the MCO problem. In practice, various scalarization techniques are often used to find effective solutions \cite{Ehrgott2005,GergelKozinov2020,Marler2004,Pardalos2017}. The present study uses the minimax convolution of partial criteria that possesses good theoretical properties
\begin{equation}
\min{f(y)} \to \min_{y\in D}{\max_{1 \leq i \leq s}{(f_i (y) \lambda_i )}} ,
\label{eq:4}
\end{equation}
where $\lambda_i \geq 0$, $1 \leq i \leq s$, $\sum_{i=1}^s{\lambda_i}=1$, are numerical indicators of the significance of each criterion.

@@

In the framework of the proposed approach the information-statistical theory of global search is used for solving the scalar optimization problems (\ref{eq:4}). For problems with several variables ($N>1$) the dimensionality reduction on the base of Peano-type mappings \cite{Sergeyev2013,Strongin2000}. While solving a problem (\ref{eq:4}) the algorithm generates a sequence of trials where the term ``trial'' means evaluation of criteria at a point of the feasible domain $D$ and computation of the convolution (\ref{eq:4}) at this point.

@@

\begin{enumerate}
  \item Obtain univariate images of all points of preceding trials mapped onto one-dimensional interval $[0,1]$ by means of Peano curves.
  \item Partition the segment $[0,1]$ into subintervals according to the images of trials performed.
  \item For each subinterval compute a numerical value called \textit{characteristic} of this subinterval.
  \item Select the subinterval with the highest characteristic.
  \item Place the point of new trial within the subinterval with the highest characteristic. 
  \item Map the new trial point to the point $y \in D$ using Peano evolvent and compute the value of the function (\ref{eq:4}) at this point. 
\end{enumerate}

@@

The second improvement is connected with a new method for calculating the characteristics of subintervals formed during univariate optimization on the internal $[0,1]$.

@@

In the second group of experiments 100 bi-criteria MCO problems have been solved. As the criteria the multiextremal functions

@@

In order to estimate efficiency of machine learning procedure in ML\_MGSA, this method tested in 2 variants: with ($\alpha>0$) and without ($\alpha=0$)
use of separating planes in the criteria space. 
Each multicriterial problem was reduced to 50 scalar problems (\ref{eq:4}) with different weight coefficients $\lambda$, distributed uniformly. Other algorithm's parameters were the same. The averaged results can be found in Table~\ref{tab:2}. \textcolor[rgb]{1,0,0}{Fig.~\ref{fig:2} additionally shows the average value of HV index change depending of the number of trials.}

@@

\begin{figure}[ht]
\center
\includegraphics[width=0.7\textwidth]{fig2.png}
\caption{Efficiency of the multicriterial algorithms compared} \label{fig:2}
\end{figure}

\textcolor[rgb]{1,0,0}{The results show that embedding the machine learning into the global search algorithm leads to }a significant reduction of the number of trials (almost 3 times) with maintaining close values of indicators HV and DU.

>>>>

\keywords{Multicriterial Problems \and Global Optimization \and Machine Learning \and Logistic Regression.}
\end{abstract}
%
%
%
\section{Introduction}
Machine learning (ML) methods are powerful tools that are applied in many areas of research.
In particular, the ML methods are successfully used for solving complicated problems of computational mathematics. One of such problem classes to which ML can be applied is the class of multicriterial optimization (MCO) models. 
In this case, ML is used, as a rule, in combination with metaheuristic algorithms \cite{Ma2014,Subraveti2019,Talbi2015} that in the case of multiextremal criteria lose in efficiency in comparison with deterministic methods \cite{Kvasov2018,Sergeyev2018}.

@@

Any effective (Pareto-optimal) variant  in which it is impossible to reduce the values of all criteria $f_i (y)$, $1 \leq i \leq s$, at once by choosing another option   can be considered as a partial solution to the MCO problem. In practice, various scalarization techniques are often used to find effective solutions \cite{Ehrgott2005,GergelKozinov2020,Marler2004,Pardalos2017}. The present study uses the minimax convolution of partial criteria that possesses good theoretical properties and allows one to find effective points by solving the problem 
\begin{equation}
\max_{1 \leq i \leq s}{ \left\{ \lambda_i f_i (y) \right\} } \rightarrow \min, \; y \in D ,
\label{eq:4}
\end{equation}
where $\lambda_i \geq 0$, $1 \leq i \leq s$, $\sum_{i=1}^s{\lambda_i}=1$, are numerical indicators (weights) of the significance of each criterion.

@@

In the framework of the proposed approach the information-statistical theory of global search is used for solving the scalar optimization problems (\ref{eq:4}). For problems with several variables ($N>1$) the dimensionality reduction on the base of Peano-type mappings \cite{Sergeyev2013,Strongin2000} is applied. While solving a problem (\ref{eq:4}) the algorithm generates a sequence of trials where the term ``trial'' means evaluation of criteria at a point of the feasible domain $D$ and computation of the convolution (\ref{eq:4}) at this point.

@@

\begin{enumerate}
  \item Obtain univariate images of all points of preceding trials mapped onto one-dimensional interval $[0,1]$ by means of Peano curves.
  \item Partition the segment $[0,1]$ into subintervals according to the images of trials performed.
  \item For each subinterval compute a numerical value called \textit{characteristic} of this subinterval.
  \item Select the subinterval with the highest characteristic.
  \item Place a point of new trial within the subinterval with the highest characteristic. 
  \item Map the new trial point to the point $y \in D$ using Peano evolvent and compute the value of the function (\ref{eq:4}) at this point. 
\end{enumerate}

@@

The second improvement is connected with a new method for calculating the characteristics of subintervals formed during univariate optimization on the interval $[0,1]$.

@@

In the second group of experiments 100 bi-criteria MCO problems have been solved. As the criteria, the multiextremal functions

@@

In order to estimate efficiency of machine learning procedure in ML\_MGSA, this method tested in 2 variants: with ($\alpha>0$) and without ($\alpha=0$)
use of separating planes in the criteria space. 
Each multicriterial problem was reduced to 50 scalar problems (\ref{eq:4}) with different weight coefficients $\lambda$, distributed uniformly. Other parameters of the algorithm were the same. The averaged results can be found in Table~\ref{tab:2}. Fig.~\ref{fig:2} additionally shows the average value of HV index depending on the number of trials.

@@

\begin{figure}[ht]
\center
\includegraphics[width=0.8\textwidth]{fig2.png}
\caption{Efficiency of the multicriterial algorithms compared} \label{fig:2}
\end{figure}

The results show that embedding the machine learning into the global search algorithm leads to a significant reduction of the number of trials (almost 3 times) with maintaining close values of indicators HV and DU.

### 2022-05-14 2022/PAVT/PAVT2022.tex

Now let us assume that $k\geq 0$ trials (in particular, $k$ can be equal to 0) are completed, and worker processes perform trials at the points $\{y(x^{k+1}), y(x^{k+2}), ...,y(x^{k+p})\}$.

Each worker process, having completed its trial at some point (without any loss of generality, let us assume this point to be $y(x^{k+1})$ corresponding to Process No 1), sends the trial result to the master process. 
In turn, the master process selects a new trial point $x^{k+p+1}$ for the worker process according to the rules described below.
Note that in this case we will have a set of preimages of the trial points
\[
I_k = \left\{ x^{k+1},x^{k+2},...,x^{k+p} \right\},
\]
at which the trials have already started, but have not yet been completed.

>>>>

Now let us assume that $k\geq 0$ trials (in particular, $k$ can be equal to 0) are completed, and worker processes perform trials at the points
$$y(x^{k+1}), \; y(x^{k+2}), ...,\; y(x^{k+p}).$$ 
Each worker process, having completed its trial at some point (without any loss of generality, let us assume this point to be $y(x^{k+1})$ corresponding to Process No 1), sends the trial result to the master process. 
In turn, the master process selects a new trial point $x^{k+p+1}$ for the worker process according to the rules described below.
Note that in this case we will have a set of preimages of the trial points
\[
I_k = \left\{ x^{k+1},x^{k+2},...,x^{k+p} \right\},
\]
at which the trials have already started, but have not yet been completed.

### 2022-05-14 2022/Sensors/kNN.tex

%Баркалов
\textcolor[rgb]{1,0,0}{Отметим тажке, что кроме space-filling curves cуществуют и другие схожие кривые, например, area-filling curves \cite{Nasso2022}. Однако широкое практическое применение получили именно Peano-Hilbert curves в силу относительной простоты построения и наличия ряда полезных свойств, основные из которых будут отмечены в Section \ref{Peano} и Section \ref{MPeano}.}

@@

The function $y^c(x)$ maps a uniform grid with step $2^{-NM}$, constructed in the interval $[0,1]$, onto a uniform grid with step $2^{-M}$, constructed in the hypercube $D$. A constructive way of establishing such a correspondence is described and theoretically justified in \cite{Strongin2000, Sergeyev2013}. 
%Баркалов
\textcolor[rgb]{1,0,0}{Алгоритмическая сложность вычисления значения функции $y^c(x)$ зависит от размерности $N$, от уровня разбиения $M$ и составляется величину $O(M\cdot N)$.}

@@

%Штанюк
\textcolor[rgb]{1,0,0}{Отметим, что метод, основанный на использовании space-filling curves, может применяться не только при работа с числовыми, но и при работе с категориальными features. Значения категориальных признаков являются дискретными, поэтому для применения данного подхода требуется сопоставить каждому дискретному значению численную величину, например, из интервала [0,1]. Конкретный способ сопоставления численных значений категориальным переменным приведен в Section \ref{RealData} (в  CarEvaluation dataset).
}%Расположение точек на отрезке [0,1] целесообразно на максимальном расстоянии друг от друга, в этом случае плотность заполняющих пространство кривых может быть уменьшена при сохранении точности обнаружения соседей.

@@

Since the hypercube $D$ from (\ref{D}) belongs to the common part of the family of hypercubes (\ref{hypercubes}) (the boundary of the hypercube $D$ is highlighted in Figure~\ref{fig2}), then, by introducing an additional function 
\[
g_0(y) = \max \{ |y_i|-2^{-1}: \; 1 \leq i \leq N \},
\] 
the original hypercube $D$ can be represented as
\[
D = \{y^l(x): x \in [0,1], \; g_0(y^l(x)) \leq 0 \}, \; 0 \leq l \leq L,
\] 
i.e. $g_0(y) \leq 0$, if $y \in D$, and $g_0(y) > 0$ if otherwise. Hence, any point $y \in D$ has its preimage $x^l \in [0,1]$ at every mapping $y^l(x), \; 0 \leq l \leq L$, i.e.
\[
y = y^l(x^l), \; x^l \in [0, 1], \; 0 \leq l \leq L.
\] 
%Баркалов
\textcolor[rgb]{1,0,0}{Алгоритмическая сложность вычисления всех прообразов $x^l \in [0, 1], \; 0 \leq l \leq L,$ зависит от размерности $N$, от уровня разбиения $M$, числа отображений $L$ и составляется величину $O(M\cdot N \cdot L)$.}

@@

%Баркалов
\textcolor[rgb]{1,0,0}{Как было сказано выше, вместо теоретической Hilbert curve $y(x)$ на практике используется evolvent $y^c(x)$, которая является приближением к $y(x)$ с точностью порядка $2^{-(M+1)}$, определяемой соотношением (\ref{prec}). При этом параметр $M$ может выбираться для каждой решаемой задачи индивидуально, исходя из точности представления исходных данных.
Например, данные в  DS-Skin dataset (Section \ref{RealData}) представляют собой координаты пикселей изображения с разрешением 256 на 256. При отображении данных на стандартный гиперкуб $D$ from (\ref{D}) расстояние по координате между двумя точками не может быть меньше $1/256$. Использование параметра $M=10$ для построения evolvent будет достаточным в этом случае. 
В случае же присвоения численных значений небольшому числу categorical features (как это было сделано для CarEvaluation dataset, see Section \ref{RealData}) расстояние по координате между объектами будет большое, значение параметра $M$ можно снизить (в экспериментах использовалось значение $M=7$).
Максимальное число одновременно используемых отображений в множестве $Y_L(x)$ from (\ref{EvSet}) также будет определяться точностью исходных данных, от которой зависит точность построения разверток $y^i(x)$.
Будем счистать, что каждое из отображений $y^i(x)$ построено с использованием одного и того же значения $M$. Тогда значение $L$ должно удовлетворять неравенству $L \leq M$. В противном случае точки $y' = y^l(x'),\; y''=y^l(x'')$ из условия (\ref{y_prime}) будут соответствовать центру одного и того же подкуба при $p>M$, т.е. будет выполняться равенство $y' = y''$  что приведет к нарушению условий Теоремы 1.}

@@

Here is a more detailed description of these algorithms.
\\[12pt]
\textit{Data preparation}
\\[12pt]
To run Algorithms A and B, we need to compute the preimages $\{x_i^0, x_i^1, ..., x_i^L\}$ on one-dimensional scales for all objects $\omega_i$, and order them in ascending order. It is convenient to use RB-tree-based data structures to store ordered one-dimensional values,
%Сысоев
\textcolor[rgb]{1,0,0}{алгоритмическая сложность операции поиска в котором имеет оценку $O(log(n))$. В качестве реализации RB-tree мы использовали контейнер set из стандартной библиотеки языка С++.}
The number of instances of tree structures is equal to the number of scales.

@@

%Сысоев
\textcolor[rgb]{1,0,0}{Следующий эксперимент проводился для определения зависимости времени поиска $K = 50$ соседей с ростом числа сэмплов для выбранной размерности $N = 4$. Число сэмплов менялось от 200 тыс. до 1 млн. Результаты, представленные на Fig.~\ref{figN}, показывают что соотношение между процедурой А и процедурой B остается примерно тем же, а выигрыш процедуры А у kNN-KD method составляет порядка 10 раз.}

@@

%Баркалов
\textcolor[rgb]{1,0,0}{Также отметим, что данный dataset включет в себя несколько категориальных атрибутов, каждому из которых соответствуют 3-4 различных значения.
В этом случае требуется выполнить сопоставление дискретным значениями численных характеристик.
В качеcтве численных значений при приведении эксперимента были выбраны внутренние узлы равномерной сетки, покрывающей отрезок [0,1]. 
Число узлов сетки выбиралось равным числу значений признаков (3 или 4, соответствевнно). Далее, i-му значению категориального признака ставилось в соответствие значение в i-м узле сетки. }

>>>>

%Баркалов
\textcolor[rgb]{1,0,0}{We should also note that in addition to space-filling curves, there are other similar curves, such as area-filling curves \cite{Nasso2022}. However, it is the Peano-Hilbert curves that are commonly used due to the relative simplicity of their construction and a number of useful properties, of which the key ones will be discussed in Section \ref{Peano} and Section \ref{MPeano}.}

@@

The function $y^c(x)$ maps a uniform grid with step $2^{-NM}$, constructed in the interval $[0,1]$, onto a uniform grid with step $2^{-M}$, constructed in the hypercube $D$. A constructive way of establishing such a correspondence is described and theoretically justified in \cite{Strongin2000, Sergeyev2013}. 
%Баркалов
\textcolor[rgb]{1,0,0}{The algorithmic complexity of computing the value of the function $y^c(x)$ depends on the dimensionality $N$, on the partitioning level $M$ and is $O(M\cdot N)$.}

@@

%Штанюк
\textcolor[rgb]{1,0,0}{Note that the method based on space-filling curves can be applied not only to numerical features but also to categorical features. The values of categorical features are discrete, so this approach requires a numerical value, e.g. from the interval [0,1], to be assigned to each discrete value.
A specific method of assigning numerical values to categorical variables is described in Section \ref{RealData} for CarEvaluation dataset.
}%Расположение точек на отрезке [0,1] целесообразно на максимальном расстоянии друг от друга, в этом случае плотность заполняющих пространство кривых может быть уменьшена при сохранении точности обнаружения соседей.

@@

Since the hypercube $D$ from (\ref{D}) belongs to the common part of the family of hypercubes (\ref{hypercubes}) (the boundary of the hypercube $D$ is highlighted in Figure~\ref{fig2}), then, by introducing an additional function 
\[
g_0(y) = \max \{ |y_i|-2^{-1}: \; 1 \leq i \leq N \},
\] 
the original hypercube $D$ can be represented as
\[
D = \{y^l(x): x \in [0,1], \; g_0(y^l(x)) \leq 0 \}, \; 0 \leq l \leq L,
\] 
i.e. $g_0(y) \leq 0$, if $y \in D$, and $g_0(y) > 0$ if otherwise. Hence, any point $y \in D$ has its preimage $x^l \in [0,1]$ at every mapping $y^l(x), \; 0 \leq l \leq L$, i.e.
\[
y = y^l(x^l), \; x^l \in [0, 1], \; 0 \leq l \leq L.
\] 
%Баркалов
\textcolor[rgb]{1,0,0}{The algorithmic complexity of computing all preimages $x^l \in [0, 1], \; 0 \leq l \leq L,$ depends on the dimensionality $N$, on the partitioning level $M$, on the number of mappings $L$ and is $O(M\cdot N \cdot L)$.}

@@

%Баркалов
\textcolor[rgb]{1,0,0}{As it was mentioned above, instead of the theoretical Hilbert curve $y(x)$, the evolvent $y^c(x)$, which is an approximation to $y(x)$ with an accuracy of the order $2^{-(M+1)}$, determined by the relation  (\ref{prec}), is used in practice. In this case, the parameter $M$ can be chosen individually for each problem to be solved, based on the accuracy of the input data representation.}

\textcolor[rgb]{1,0,0}{For example, the data in the DS-Skin dataset (Section \ref{RealData}) are pixel coordinates of an image with a resolution of 256 by 256. When mapping data to the standard hypercube $D$ from (\ref{D}), the coordinate distance between two points cannot be less than $1/256$. In this case, using the parameter $M=10$ to construct the evolvent will be sufficient.
If we assign numerical values to a small number of categorical features (as was done for the CarEvaluation dataset, see Section \ref{RealData}), the  coordinate distance between the features will be large, so the value of $M$ can be reduced (in our experiments, we used $M=7$).}

\textcolor[rgb]{1,0,0}{The maximum number of simultaneous mappings in the set $Y_L(x)$ from (\ref{EvSet}) will also be determined by the accuracy of the input data, which affects the accuracy of constructing the evolvents $y^i(x)$.
Let us assume that each of the mappings  $y^i(x)$ is constructed using the same value of $M$. Then the value of $L$ must satisfy the inequality $L \leq M$. Otherwise the points $y' = y^l(x'),\; y''=y^l(x'')$ from condition (\ref{y_prime}) will correspond to the centre of the same subcube with $p>M$, i.e. the equality  $y' = y''$ will be fulfilled, leading to violation of the conditions of Theorem 1.}

@@

Here is a more detailed description of these algorithms.
\\[12pt]
\textit{Data preparation}
\\[12pt]
To run Algorithms A and B, we need to compute the preimages $\{x_i^0, x_i^1, ..., x_i^L\}$ on one-dimensional scales for all objects $\omega_i$, and order them in ascending order. It is convenient to use RB-tree-based data structures to store ordered one-dimensional values,
%Сысоев
\textcolor[rgb]{1,0,0}{where the algorithmic complexity of the search operation has the estimate $O(log(n))$. We used the container set from the standard C++ library as an implementation of the RB-tree.}
The number of instances of tree structures is equal to the number of scales.

@@

%Сысоев
\textcolor[rgb]{1,0,0}{The following experiment was conducted to determine the dependence of the search time for $K = 50$ neighbours with increasing number of samples for the chosen dimensionality $N = 4$. The number of samples was varied from 200 thousand to 1 million. The results presented in Figure~\ref{figN} show that the difference between procedure A and procedure B remains approximately the same, and the gain of procedure A over the kNN-KD method is about 10 times.}

@@

%Баркалов
\textcolor[rgb]{1,0,0}{Also note that this dataset will include several categorical attributes, each corresponding to 3-4 different values.
In this case, it is necessary to assign numerical characteristics to discrete values.
The numerical values used in the experiment were the internal nodes of a uniform grid covering the interval $[0,1]$. 
The number of grid nodes was chosen to be equal to the number of feature values (3 or 4, respectively). Then, the $i$-th value of the categorical feature was assigned to the $i$-th node of the grid.}
